Characteristic polynomials of p-adic matrices

### 2017-04-18 charpoly/charpoly.tex
Rewrote abstract and "The contribution of this paper"

\begin{abstract}
We analyze the precision of the characteristic polynomial $\chi(M)$ of
an $n \times n$ $p$-adic matrix $M$ using differential precision methods
developed previously.  When $M$ is integral with precision $O(p^N)$,
we give a criterion (checkable in time $\softO(n^\omega)$) for
$\chi(M)$ to have precision exactly $O(p^N)$.  We also give a $\softO(n^3)$
algorithm for determining the optimal precision when the criterion is not
satisfied, and give examples when the precision is larger than $O(p^N)$.
\end{abstract}

@@

Thanks to the application of the framework of differential precision
of \cite{caruso-roe-vaccon:14a} in \cite{caruso-roe-vaccon:15a},
we know that the precision of the characteristic polynomial
$\chi_M$ of a matrix $M \in M_n(\mathbb{Q}_p)$
is determined by the adjugate $\adj(X -M).$
In this article, we provide:
\begin{enumerate}
\renewcommand{\itemsep}{0pt}
\item Proposition \ref{prop:shortcom}: a factorization of $\adj(X -M)$
as a product of two rank-$1$ matrices (when $M$ has a cyclic vector),
computable in $\softO(n^\omega)$ operations by Theorem \ref{thm:compute_shortcom}
\item Corollary \ref{cor:prec_gain}: a simple, $\softO(n^\omega)$ criterion
to decide whether $\chi_M$ is defined at precision
higher than the precision of $M$ (when $M \in M_n(\mathbb{Z}_p)$).
\item Theorem \ref{thm:alg3}: a $\softO(n^3)$ algorithm with operations in $\Zp$
to compute the optimal precision on each coefficient of $\chi_M$ (when $M$
is given with uniform precision on its entries).
\item Proposition \ref{prop:opteigenvalues}: a $\softO(n^\omega)$
algorithm to compute the optimal precision on each eigenvalue of $M$
\end{enumerate}

>>>>

\begin{abstract}
We analyze the precision of the characteristic polynomial $\chi(M)$ of 
an $n \times n$ $p$-adic matrix $M$ using differential precision methods 
developed previously.  When $M$ is an integral matrix whose entries are 
all given at the same precision $O(p^N)$, we give a criterion (checkable 
within $\softO(n^\omega)$ operations in $\Fp$) for the existence of a
coefficient of $\chi(M)$ with more accuracy that $O(p^N)$.
In general, we provide two algorithms for determining the optimal 
precision on the coefficients of $\chi(M)$ and on their eigenvalues.
We provide evidences showing that classical algorithms do not reach
this optimal precision in general.
\end{abstract}

@@

This paper provides a theoretical and concrete study of the propagation 
of the precision during the computation of the characteristic polynomial 
of matrix defined over the $p$-adics. Thanks to the application of the 
general framework of differential precision developed in 
\cite{caruso-roe-vaccon:14a,caruso-roe-vaccon:15a}, we know that the 
optimal precision of the characteristic polynomial $\chi_M$ of a matrix 
$M \in M_n(\mathbb{Q}_p)$ is controled by the adjugate $\adj(X{-}M).$
In this article, we provide:
\begin{enumerate}
\renewcommand{\itemsep}{0pt}
\item Proposition \ref{prop:shortcom}: a short representation of
the matrix $\adj(X{-}M)$ whose size is $O(n^2)$ elements of the base
field (instead to $O(n^3)$ for a dense representation);
\item two algorithms to compute $\adj(X{-}M)$: the first one is 
division-free and performs $\softO(n^3)$ operations in $\Zp$ while 
the second one may perform divisions and has complexity $\tilde 
O(n^\omega)$;
\end{enumerate}
and deduce from this the following applications to the question
we are interested in:
\begin{enumerate}
\renewcommand{\itemsep}{0pt}
\setcounter{enumi}{2}
\item Corollary \ref{cor:prec_gain}: a simple criterion to decide 
whether $\chi_M$ is defined at precision higher than the precision of 
$M$ (when $M$ lies in $M_n(\mathbb{Z}_p)$ and all its entries are given 
at the same precision);
\item two algorithms to compute the optimal precision on each 
coefficient of the characteristic polynomial of a matrix $M \in 
M_n(\Qp)$ (whose entries may be given at different precision): the first 
one is division-free and has complexity $\softO(n^3)$ while the second 
one may perform divisions but has complexity $\softO(n^\omega)$;
\item Proposition \ref{prop:opteigenvalues}: a $\softO(n^\omega)$
algorithm to compute the optimal precision on each eigenvalue of $M$.
\end{enumerate}
Moreover, we provide evidences showing that classical algorithms for 
computing the characteristic polynomials do not reach the optimal 
precision on average.
